Genetic draft and the evolution of ``irreducible complexity"

### 2017-09-15 valley_crossing.tex
sketching some heuristics, filling in theory

>>>>

\section*{Model}

Our model features a population of $N$ haploid individuals in which an effectively infinite number of ``background" loci, with weak fitness effects, are currently segregating.
We focus on two loci, which are initially fixed for alleles $a$ and $b$.
These alleles mutate to $A$ and $B$ at rates $\mu_a$ and $\mu_b$, respectively (we neglect back mutations).
Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_a$, $1-\delta_b$, and $1+s$ respectively, with $\delta \geq 0$ (so the intermediate genotypes are deleterious) and $1/N < s < 1$ (so the complex adaptation is strongly beneficial).
For the remainder of our analysis, we will assume that $\mu_a = \mu_b$ and $\delta_a = \delta_b$: we discuss the effects of relaxing this assumption in the appendix.
We further assume that the large number of loci of weak effect contribute to a background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate at which the mean fitness advances.
Organisms cross over at rate $\rho$ per site per generation.
Our goal is to compute the expected time $\mathbb{E}\left[ T\right]$ for such a complex adaptation to arise and fix in the population, as a function of the parameters.

In the case where $\rho = 0$ and the fitness variance $\sigma^2$ is small ($N\sigma \ll 1$), the rate of valley crossing agrees with Weissman et al. (2009): when $\rho > 0$, the rate agrees with Weissman et al. (2010).
This is because, provided that $N\sigma \ll 1$, genetic drift is the major evolutionary process affecting the frequency of deleterious intermediates, and the fixation probability of the complex adaptation simply scales with $s$.
We turn to the regime where $N\sigma \gg 1$, meaning that draft is more important than drift, but the fixation probability of the full complex adaptation is suppressed.
By the central limit theorem, the background loci create a background fitness wave whose shape is roughly Gaussian.
We assume that the constant influx of weak effect beneficial mutations keeps the fitness variance $\sigma^2$ roughly constant.
We sketch some heuristics in this limit before providing a more formal analysis.

First, we consider the case where $\rho = 0$ and focus on what Weissman et al. (2009) refer to as the process of ``stochastic tunneling", where $N\mu \ll 1$.
As we shall see, the results in the $\rho = 0$ and low $\rho$ limits can easily be extended to sexually recombining populations: moreover, understanding the rate of tunneling will allow us to easily examine the rate at which sequential fixation and deterministic fixation occur.
In stochastic tunneling, a deleterious intermediate arises (at rate $N\mu$), and its lineage persists over a short time $T$, forming a transient ``bubble".
If the number of individuals in the lineage is given by $n(t)$, then the total number of individuals in the ``bubble" is $w = \int_{t_0}^T n(t) dt$.
We refer to $w$ as the ``weight" of the bubble.
The overall rate of valley crossing becomes $1/N\mu p$, with $p$ the probability that a particular bubble will give rise to a successful double mutant.
Thus, given that a deleterious intermediate appears, it has $w$ ``chances" to generate a successful complex adaptation, which it does at rate $\sim \mu s$: double mutants arise at rate $\mu$ and have a probability $\sim s$ of fixing.
The probability of achieving a weight of at least $w$ scales as $1/\sqrt{w}$ (or, equivalently, $P(w) \sim w^{-3/2}$), meaning that with probability $\sqrt{\mu s}$, a bubble gives rise to a successful single mutant.
This probability drops off rapidly at $1/\delta^2$, so for higher values of $\delta$ (so called ``deleterious tunneling"), the rate at which new complex adaptations are generated starts to fall: the size of the mutant subpopulation is bounded from above by roughly $1/\delta$.

In the high $N\sigma$ regime, the argument is much the same, but the underlying dynamics are different.
In stark contrast to the drift case, the fate of a complex adaptation depends strongly on the fitness $x$ of the genetic background on which it appears.
Individuals with fitness lower than some critical fitness $x_c$ are almost guaranteed to go extinct.
This means the fixation probability of the complex adaptation no longer scales with $s$, and $P(w) \sim w^{-2}$ instead of $w^{-3/2}$, meaning that the probability of reaching weight $w$ is $\sim 1/w$.
However, individuals with higher fitnesses can be destined for fixation, even if they are carrying a significant load of deleterious mutations.
This means the dependence on $\delta$ is significantly lessened: even deep fitness valleys can be crossed by a rapid enough wave.
(Still needed: expression for fixation probability of beneficial mutant--this should be easy enough to obtain.
Can start with equation 3 of Neher and Shraiman (2009) and drop the recombination term.
This is tantamount to a ``mean field" approximation but might still work out okay.
Other possible approach is to use something like Desai and Fisher's (2007) expression for the fixation probability and rescale it to a traveling wave model.
Once we have the fixation probability, studying sequential and deterministic fixation will be easier than studying tunneling, since sequential fixation is essentially two different fixation events.)

We now turn to the case where $\rho > 0$.
Because we are explicitly dealing with crossovers, the results of Neher, Shraiman, and Kessinger (2013) are useful here.
We consider the genome as divided into a series of effectively asexual ``blocks" in which the recombination rate is very low.
The size of each block is determined by a balance between recombination, which chops up the blocks, and the rate of adaptation, which amplifies them.
The size of such a block is not a parameter of our model but is given by $\xi_b = \frac{\sigma^2}{2L\rho^2 c \log N\sigma_b}$, with $L$ the number of loci, $c$ a constant of order $1$, and $\sigma_b$ the proportion of the fitness variance segregating within the block.
If $\xi_b$ is large enough that both focal loci fall within the same ``block", then an analysis similar to that of Neher and Shraiman (2009) may be appropriate: the two loci effectively segregate within the same traveling wave.
If not, then the value of $N\sigma_b$ becomes the relevant factor.
$N\sigma_b \gg 1$ implies that each locus effectively operates within its own traveling wave: in order for the full complex adaptation to arise, establish, and fix, both must be at high enough frequency at the same time for either mutation or recombination to produce a double mutant.
On the other hand, $N\sigma_b \ll 1$ implies that each locus effectively evolves on a neutral genetic background neutrally, and we recover the dynamics of genetic drift.
